Reinforcement \change{L}earning for \change{O}ptions\\on \change{T}arget \change{V}olatility \change{F}unds\thanks{Supported by ISCRA from Cineca, \email{iscra@cineca.it}.}

### 2023-07-04 tex/m2ef/daluiso_nastasi_pallavicini_polo_rl_for_options_on_tvf.tex
paper final version

	\subsection{The strategy price process}
	We work on a filtered probability space $\left(\Omega, \mathcal{F}, \{\mathcal{F}_t\}_{t\geq0}, \mathbb{P}  \right)$ satisfying the usual assumptions for a market model, where $\mathbb{P}$ is the physical probability measure \remove{representing the actual distribution of supply and demand shocks on equities prices}.

@@

	It is possible to derive the expression for $\omega_t$ assuming a generic continuous semi-martingales dynamics under the risk-neutral measure for the underlying securities, so that Equation~\eqref{eq:risk_neutral} can be written as 
	\begin{equation}
		\frac{dS_t^i}{S_t^i} = \left(r_t - \mu_t^i \right)dt + \nu_t^i \cdot dW_t \; ,
	\label{eq:Equity_process}\end{equation}
	where $\nu_t$ is an adapted matrix process ensuring the existence of a solution for the stochastic differential equation \remove{(SDE)} and $W_t$ is a $n$-dimensional vector of Brownian motions under $\mathbb{Q}$. Under these assumptions we can derive an expression for $\omega_t$, and we get\footnote{In all formulae the norm for a vector $a$ is defined as $\|a\|\coloneqq\sqrt{a\cdot a}$.}
	\begin{equation}
		\omega_t = \frac{\Bar{\sigma}}{\norm{ \alpha_t\cdot \nu_t }} \; .
	\end{equation}
	Hence, putting this last result in the dynamics of $I_t$ we obtain

@@

	\subsection{Stochastic optimal control problem}
	In presence of valuation adjustments, we need to solve the full optimization problem. We discretize the optimal strategy $\alpha_t$ as\footnote{We use the symbol $\mathbf{1}_A$ for the indicator function of a subset $A$.}
	\begin{equation}
		\alpha_t \coloneqq \sum_k \mathbf{1}_{ \{t \in [T_{k-1}, T_k)\}}\alpha_{T_{k-1}} \; ,
	\label{eq:piecewise_strategy}\end{equation}
	according to a time grid $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq t$ the pricing date and $T_m \coloneqq T$ the maturity of the option. 
	Therefore we can apply the dynamic programming principle to express the optimal $\alpha_t$ at time $T_{k-1}$ as
	\begin{equation}
		\alpha_{T_{k-1}}\coloneqq \argmax_{\alpha} \left\{\mathbb{E}_{T_{k-1}}\left[D\left(T_{k-1},T_{k} \right) V_{T_{k}}\left(X_{T_{k}}, I_{T_{k}}(\alpha)\right) \mid X_{T_{k-1}}, I_{T_{k-1}}\right]\right\} \; ,
	\label{eq:recursion}\end{equation}
	where $V_{T_k}$ is the option value at time $T_k$ and $X$ is any Markovian state such that the drift and the diffusion coefficient of $I_t$ are a function of $\left(X_t,I_t,\alpha_t\right)$ and we indicated with $I_{T_k}\left(\alpha\right)$ the value of the strategy at time $T_{k}$ for a given choice of the weights $\alpha$ at time $T_{k-1}$. We calculate the strategy value for the optimal weights $\alpha_{T_{k-1}}$, namely $I_{T_k}\left(\alpha_{T_{k-1}}\right)$ \remove{for any given strategy $\alpha_{T_{k-1}}$} by a suitable discretization of \eqref{eq:TVS_last} starting from $X_{T_{k-1}}$ and $I_{T_{k-1}}$.
	Thus, once collected the elements $\{X_{T_{k}}, I_{T_{k}}\left(\alpha_{T_{k-1}}\right)\}$, the derivative price is given by:
	\begin{equation}
	\begin{aligned}
			V_{T_{k-1}}\left(X_{T_{k-1}}, I_{T_{k-1}}\right)= \\ 
			\mathbb{E}_{T_{k-1}}\left[D\left(T_{k-1},T_{k}\right) V_{T_{k}}\left(X_{T_{k}}, I_{T_{k}}\left(\alpha_{T_{k-1}}\right)\right) \mid X_{T_{k-1}}, I_{T_{k-1}}\right] \; ,
	\end{aligned}
	\label{eq:recursion2}\end{equation}
	while the iteration starts from maturity date where the boundary condition is set equal to the payoff function:
	\begin{equation}
		V_{T_m} = \Phi\left(I_{T_m}\right) \; .
	\end{equation}

@@

	\subsubsection{Unconstrained allocation strategy: closed form solution}
	In absence of constraints on the allocation strategy, we are able to derive a closed form solution to the BS problem \eqref{eq:BS_optimal_strategy}.
	\begin{lemma}
		Let \remove{be} $\mu, \alpha \in \mathbb{R}^n$, $\nu \in \mathbb{R}^{n\times n}$ be a full rank matrix and $\Sigma\coloneqq \nu \tr{\nu}$. Then the closed solution of the optimization problem \eqref{eq:BS_optimal_strategy} is 
		\begin{equation}
				\alpha^* = - \frac{\Sigma^{-1} \cdot \mu}{\norm{(\Sigma^{-1} \cdot \mu)\cdot \nu}} \; .
		\end{equation}
	\end{lemma}
	\begin{proof}
	Since the argument of the minimum \eqref{eq:BS_optimal_strategy} is zero-homogeneous, then we can rewrite the problem as
	\begin{equation}
			\begin{aligned}
				&\text { minimize } \alpha \cdot \mu\\
				&\text { subject to } \|\alpha \cdot \nu\|^2=1 \; .
		\end{aligned}
	\end{equation}
	By setting the Lagrangian function associated with the problem
	\begin{equation}
		\mathcal{L}\left(\alpha, \lambda\right)=\alpha \cdot \mu-\lambda\left(\|\alpha \cdot \nu\|^2-1\right) \; ,
	\end{equation}
	we obtain the first order conditions
	\begin{eqsys}
		\frac{\partial \mathcal{L}}{\partial \alpha}=\mu -2\lambda \Sigma \cdot \alpha=0 \\
		\frac{\partial \mathcal{L}}{\partial \lambda}= \norm{\alpha \cdot \nu}^2-1=0
		\; ,
	\end{eqsys}
	Then, by applying simple algebra, we obtain the analytical form for the free optimal strategy
	\begin{equation}
		\alpha^* = \pm \frac{\Sigma^{-1} \cdot \mu}{\norm{(\Sigma^{-1} \cdot \mu)\cdot \nu}} \; .
	\label{eq:analytical_solution_free_strategy}\end{equation}
	We take the minus sign to get the minimum value of the TVS local drift while the plus sign for the maximum one (put payoff). $\Box$
	\end{proof}

@@

	\subsection{Hamilton-Jacobi-Bellman equation for target volatility options}\label{subsec:HJB}
	In this section, we \remove{want to} provide to the reader a formal description of the dynamic problem associated with options on target volatility strategies by writing the Hamilton-Jacobi-Bellman (HJB) equation for the derivative price. We prove that from this equation we can recover the same closed formula \eqref{eq:BS_optimal_strategy} derived above from the Gy\"ongy Lemma for the time-dependent BS model.

@@

	Once the agent has selected the action $a_k$ sampled from the current policy, it receives at the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment. In this algorithm we have defined the reward function as follows
	\begin{equation}
		\change{r_{k+1}\change{\coloneqq}\left\{\begin{array}{ll}
			D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
			0 & \text { otherwise }
		\end{array}\right.
	\label{eq:reward1} \; .} \end{equation}

@@

	In this algorithm, we parameterize the agent policy with a feed forward neural network (FFNN), such that $\theta_j$ coincides with the hidden weights, $s_k$ with the input neurons, and $a_k$ with the output \change{ones}. In this way, we are dealing with a deterministic policy, \change{that maps each state directly to a single action without any additional randomness or probability distribution;} \remove{where its} \change{the map} functional form is given by the neural network. The parameters update is performed as follows: the agent collects a finite batch of experiences interacting with the environment in a set of episodes $\tau$, then the loss-\change{value} function is evaluated as
	\begin{equation}
	L(\theta_j) =\hat{\mathbb{E}}\left[ \sum_{k=0}^{m-1} r_{k+1} | \theta_j \right] \; ,
	\label{eq:direct_policy_loss}\end{equation}
	\begin{equation}
	\hcancel[red]{\change{L(\theta_j) = D(0,T;\zeta)\hat{\mathbb{E}}\left[ \Phi(I_T(\alpha(\theta_j))) \right]}} \; ,
	\end{equation}  
	where the expectation $\hat{\mathbb{E}}[\dots]$ indicates the empirical average on the batch. Then the parameters are updated by plugging the policy into a stochastic gradient ascent algorithm \change{\cite{Sutton2018} with the aim of maximizing Equation~\eqref{eq:direct_policy_loss}}. \remove{The choice of Equation~\eqref{eq:direct_policy_loss} is justified by the fact that, when the algorithm will find the optimal set $\theta^*$, then $L(\theta^*)$ will be a good proxy of the optimal option price.}

@@

	\subsection{Proximal policy optimization approach}
	As for the direct policy \change{search} approach, we model the pricing problem considering an episode that takes place on the time-grid $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$. Again we choose as observation state the block defined in Equation~\eqref{eq:state} and the agent policy coincides with the risky allocation strategy $\alpha$ \eqref{eq:action}. \remove{An important difference from the previous method is that: once the agent has selected the action $a_k$ sampled from the current policy, it receives at the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment.}

	\remove{In our work, we have defined two different reward functions with the purpose to analyze which could help the agent to learn more efficiently the optimal policy $\pi^* = \pi^{\theta^*}$. 
	Our first definition is} 
	\begin{equation}
		\hcancel[red]{\change{r_{k+1} =\left\{\begin{array}{ll}
			D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
			0 & \text { otherwise }
		\end{array}\right.}}
	\; . \end{equation}
	\remove{Therefore, during the whole episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the episode, increasing the probability to obtain a slow learning. However, if the agent has learnt $\pi^*$, the average cumulative reward per episode will coincide with the optimal TVO price.}

	\change{As the PPO algorithm is more sophisticated than the previous one, we have decided to test its performance by means of two different reward functions. The former is the same as defined by Equation~\eqref{eq:reward1}, while the latter is given by}
	\remove{The second reward function we have defined is}
	\begin{equation}
		r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] \; ,
	\label{eq:reward2}\end{equation}
	where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is a proxy of the residual option price defined by
	\begin{equation}
	\begin{aligned}
		V_{\text{BS}}(T_k) &\coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}},K),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \; ,\\
		V_{\text{BS}}(T_0)&=0 \; , 
	\end{aligned}
	\end{equation} 
	with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} evaluated at the state $s_k$. In this form the agent actions are hidden inside the term $I_t$ used to compute the TVS forward curve $F^{\text{TVS}}$ defined by Equation~\eqref{eq:tvs_forward}. In this case, the reward function does not suffer of nil values for $0<T_k<T$: the RL agent always gets a feedback from the environment for its choices.  \remove{The} \change{Notice that the} hyper-parameter $\gamma$ plays the role of a discount factor in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future.
	If we take the cumulative reward per episode and set the PPO parameters\footnote{We refer to \cite{Schulman2016} for a more detailed description for the generalized advantage estimation parameter $\lambda$.} $\gamma =\lambda =1$ we obtain
	\begin{equation}
		R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma=\lambda =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+ \; ,
	\end{equation} 
	which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we conjecture that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

	In the \textsc{OpenAI Baselines} implementation of the PPO, the agent policy is parameterized again by a neural network; as for the previous method we have chosen an FFNN. \remove{In particular with PPO we deal with a stochastic policy, whose functional form is} \change{More precisely, training is based on a stochastic policy, i.e. the action is drawn from a probability law, and the neural network is used to determine its parameters. In settings like ours where the action has a continuous domain, such law is} a multivariate diagonal Gaussian distribution where the mean $\mu^\theta(s_k)$ is the output vector of the FFNN and the log-standard deviation $\log\sigma$ is an \remove{external} \change{additional trainable} parameter:
	\begin{equation}
		\pi^\theta(s_k) \sim \mathcal{N}(\mu^\theta(s_k),e^{\log\sigma}) \; .
	\end{equation}
	As one can observe, $\log\sigma$ is state-independent, but it is reduced as the number of the PPO update iterations increases. The idea is that the log-standard deviation will be higher at the beginning of the training phase in order to guarantee a good exploration of the action space while it will be lower at the end to avoid too much noise in the proximity of the optimal policy. \change{Eventually at the end of training, the standard deviation is put to zero, so that the action becomes deterministically equal to the output of the FFNN.}

@@

	\section{Numerical investigations}\label{sec:Numerical_investigations}
	Here we present the numerical results obtained with our proposed methods. We focus our analysis on a European call option on a TVS with the following product details
	\begin{equation}\nonumber
		I_0 = K = 1 \, \text{EUR} \; , \quad T=2 \, \text{yr} \; , \quad \bar{\sigma} = 5\% \; \change{,}
	\end{equation}
	\change{where $I_0$ is the starting value of the TVS-based portfolio.}
	Moreover\change{,} without loss of generality\change{,} we consider the case of a\change{n unconstrained} \remove{completely free} allocation strategy $\alpha$. The extension to the constrained case is easy.

@@

	Although in Section~\ref{subsec:BS_model} we have proved that under the Black and Scholes model the Equation~\eqref{eq:BS_optimal_strategy} solves the control problem, we want to take advantage of this \textit{a priori} solution as a benchmark to gather evidence on the robustness of our \remove{Reinforcement Learning} \change{RL} approach and to check if our analytical result is correct. Moreover, we use the BS model as numerical laboratory to perform fine-tuning tests for the RL algorithms hyper-parameters and analyze how they impact the final results and performances.

	We recall that in both the algorithms we parameterize the agent policy with an FFNN; thus this is completely characterized by the following hyper-parameters: number of hidden layers, number of neurons per hidden layer, \change{and} activation function per hidden layer. This is due to the fact that in this \remove{Reinforcement Learning} \change{RL} problem the number of the input neurons is equal to the state space dimension \eqref{eq:state}, while the number of the output ones coincides with the action space dimension \eqref{eq:action}. It is well known in the literature that neural networks give better performances in the training phase if the input data are well normalized \cite{Puheim2014}, \cite{Sola1997}. Thus we choose as state $s_k$ the following normalized block  
	\begin{equation}
		[\log(S_{T_k}/F(0,T_k)), I_{T_k}/I_0, T_k ] \quad \forall T_k\in\mathcal{T} \; ,
	\label{eq:input_state}\end{equation}
	where $F(t,T)$ is the forward curve vector of the risky assets from $t$ to $T$. In Equation~\eqref{eq:input_state} we have chosen as Markovian state $X_{T_k}$ the martingale term of the securities dynamics. In this way\change{,} we have that in the input block the variables have the same order of magnitude.

	\subsection{Black and Scholes: hyper-parameters fine tuning}
	We use the BS environment as \change{a} toy model to understand which parameters of the RL algorithm play key roles in the training and testing phase. 
	Our first approach has been the direct policy one since it represents a natural way to tackle the problem: since our goal is to find the allocation 
	strategy that maximizes the option price, we update the FFNN weights following the gradient direction of the loss function defined in Equation~\eqref{eq:direct_policy_loss}.

>>>>

	\subsection{The strategy price process}
	We work on a filtered probability space $\left(\Omega, \mathcal{F}, \{\mathcal{F}_t\}_{t\geq0}, \mathbb{P}  \right)$ satisfying the usual assumptions for a market model, where $\mathbb{P}$ is the physical probability measure.

@@

	It is possible to derive the expression for $\omega_t$ assuming a generic continuous semi-martingales dynamics under the risk-neutral measure for the underlying securities, so that Equation~\eqref{eq:risk_neutral} can be written as 
	\begin{equation}
		\frac{dS_t^i}{S_t^i} = \left(r_t - \mu_t^i \right)dt + \nu_t^i \cdot dW_t \; ,
	\label{eq:Equity_process}\end{equation}
	where $\nu_t$ is an adapted matrix process ensuring the existence of a solution for the stochastic differential equation and $W_t$ is a $n$-dimensional vector of Brownian motions under $\mathbb{Q}$. Under these assumptions we can derive an expression for $\omega_t$, and we get\footnote{In all formulae the norm for a vector $a$ is defined as $\|a\|\coloneqq\sqrt{a\cdot a}$.}
	\begin{equation}
		\omega_t = \frac{\Bar{\sigma}}{\norm{ \alpha_t\cdot \nu_t }} \; .
	\end{equation}
	Hence, putting this last result in the dynamics of $I_t$ we obtain

@@

	\subsection{Stochastic optimal control problem}
	In presence of valuation adjustments, we need to solve the full optimization problem. We discretize the optimal strategy $\alpha_t$ as\footnote{We use the symbol $\mathbf{1}_A$ for the indicator function of a subset $A$.}
	\begin{equation}
		\alpha_t \coloneqq \sum_k \mathbf{1}_{ \{t \in [T_{k-1}, T_k)\}}\alpha_{T_{k-1}} \; ,
	\label{eq:piecewise_strategy}\end{equation}
	according to a time grid $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq t$ the pricing date and $T_m \coloneqq T$ the maturity of the option. 
	Therefore we can apply the dynamic programming principle to express the optimal $\alpha_t$ at time $T_{k-1}$ as
	\begin{equation}
		\alpha_{T_{k-1}}\coloneqq \argmax_{\alpha} \left\{\mathbb{E}_{T_{k-1}}\left[D\left(T_{k-1},T_{k} \right) V_{T_{k}}\left(X_{T_{k}}, I_{T_{k}}(\alpha)\right) \mid X_{T_{k-1}}, I_{T_{k-1}}\right]\right\} \; ,
	\label{eq:recursion}\end{equation}
	where $V_{T_k}$ is the option value at time $T_k$ and $X$ is any Markovian state such that the drift and the diffusion coefficient of $I_t$ are a function of $\left(X_t,I_t,\alpha_t\right)$ and we indicated with $I_{T_k}\left(\alpha\right)$ the value of the strategy at time $T_{k}$ for a given choice of the weights $\alpha$ at time $T_{k-1}$. We calculate the strategy value for the optimal weights $\alpha_{T_{k-1}}$, namely $I_{T_k}\left(\alpha_{T_{k-1}}\right)$ by a suitable discretization of \eqref{eq:TVS_last} starting from $X_{T_{k-1}}$ and $I_{T_{k-1}}$.
	Thus, once collected the elements $\{X_{T_{k}}, I_{T_{k}}\left(\alpha_{T_{k-1}}\right)\}$, the derivative price is given by:
	\begin{equation}
	\begin{aligned}
			V_{T_{k-1}}\left(X_{T_{k-1}}, I_{T_{k-1}}\right)= \\ 
			\mathbb{E}_{T_{k-1}}\left[D\left(T_{k-1},T_{k}\right) V_{T_{k}}\left(X_{T_{k}}, I_{T_{k}}\left(\alpha_{T_{k-1}}\right)\right) \mid X_{T_{k-1}}, I_{T_{k-1}}\right] \; ,
	\end{aligned}
	\label{eq:recursion2}\end{equation}
	while the iteration starts from maturity date where the boundary condition is set equal to the payoff function:
	\begin{equation}
		V_{T_m} = \Phi\left(I_{T_m}\right) \; .
	\end{equation}

@@

	\subsubsection{Unconstrained allocation strategy: closed form solution}
	In absence of constraints on the allocation strategy, we are able to derive a closed form solution to the BS problem \eqref{eq:BS_optimal_strategy}.
	\begin{lemma}
		Let $\mu, \alpha \in \mathbb{R}^n$, $\nu \in \mathbb{R}^{n\times n}$ be a full rank matrix and $\Sigma\coloneqq \nu \tr{\nu}$. Then the closed solution of the optimization problem \eqref{eq:BS_optimal_strategy} is 
		\begin{equation}
				\alpha^* = - \frac{\Sigma^{-1} \cdot \mu}{\norm{(\Sigma^{-1} \cdot \mu)\cdot \nu}} \; .
		\end{equation}
	\end{lemma}
	\begin{proof}
	Since the argument of the minimum \eqref{eq:BS_optimal_strategy} is zero-homogeneous, then we can rewrite the problem as
	\begin{equation}
			\begin{aligned}
				&\text { minimize } \alpha \cdot \mu\\
				&\text { subject to } \|\alpha \cdot \nu\|^2=1 \; .
		\end{aligned}
	\end{equation}
	By setting the Lagrangian function associated with the problem
	\begin{equation}
		\mathcal{L}\left(\alpha, \lambda\right)=\alpha \cdot \mu-\lambda\left(\|\alpha \cdot \nu\|^2-1\right) \; ,
	\end{equation}
	we obtain the first order conditions
	\begin{eqsys}
		\frac{\partial \mathcal{L}}{\partial \alpha}=\mu -2\lambda \Sigma \cdot \alpha=0 \\
		\frac{\partial \mathcal{L}}{\partial \lambda}= \norm{\alpha \cdot \nu}^2-1=0
		\; ,
	\end{eqsys}
	Then, by applying simple algebra, we obtain the analytical form for the free optimal strategy
	\begin{equation}
		\alpha^* = \pm \frac{\Sigma^{-1} \cdot \mu}{\norm{(\Sigma^{-1} \cdot \mu)\cdot \nu}} \; .
	\label{eq:analytical_solution_free_strategy}\end{equation}
	We take the minus sign to get the minimum value of the TVS local drift while the plus sign for the maximum one (put payoff). $\Box$
	\end{proof}

@@

	\subsection{Hamilton-Jacobi-Bellman equation for target volatility options}\label{subsec:HJB}
	In this section, we provide to the reader a formal description of the dynamic problem associated with options on target volatility strategies by writing the Hamilton-Jacobi-Bellman (HJB) equation for the derivative price. We prove that from this equation we can recover the same closed formula \eqref{eq:BS_optimal_strategy} derived above from the Gy\"ongy Lemma for the time-dependent BS model.

@@

	Once the agent has selected the action $a_k$ sampled from the current policy, it receives at the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment. In this algorithm we have defined the reward function as follows
	\begin{equation}
		r_{k+1}\coloneqq\left\{\begin{array}{ll}
			D(0,T;\zeta)(I_T(\alpha(\theta_j)) - K)^+ & \text { if } T_{k+1} = T \\
			0 & \text { otherwise }
		\end{array}\right.
	\label{eq:reward1} \; . \end{equation}

@@

	In this algorithm, we parameterize the agent policy with a feed forward neural network (FFNN), such that $\theta_j$ coincides with the hidden weights, $s_k$ with the input neurons, and $a_k$ with the output ones. In this way, we are dealing with a deterministic policy, that maps each state directly to a single action without any additional randomness or probability distribution; the map functional form is given by the neural network. The parameters update is performed as follows: the agent collects a finite batch of experiences interacting with the environment in a set of episodes $\tau$, then the loss-value function is evaluated as
	\begin{equation}
	L(\theta_j) =\hat{\mathbb{E}}\left[ \sum_{k=0}^{m-1} r_{k+1} | \theta_j \right] \; ,
	\label{eq:direct_policy_loss}\end{equation}
	where the expectation $\hat{\mathbb{E}}[\dots]$ indicates the empirical average on the batch. Then the parameters are updated by plugging the policy into a stochastic gradient ascent algorithm \cite{Sutton2018} with the aim of maximizing Equation~\eqref{eq:direct_policy_loss}.

@@

	\subsection{Proximal policy optimization approach}
	As for the direct policy search approach, we model the pricing problem considering an episode that takes place on the time-grid $\mathcal{T}\coloneqq \{T_0,...,T_k,...,T_m\}$ with $T_0\coloneqq 0$ and $T_m \coloneqq T$. Again we choose as observation state the block defined in Equation~\eqref{eq:state} and the agent policy coincides with the risky allocation strategy $\alpha$ \eqref{eq:action}. 
	As the PPO algorithm is more sophisticated than the previous one, we have decided to test its performance by means of two different reward functions. The former is the same as defined by Equation~\eqref{eq:reward1}, while the latter is given by
	\begin{equation}
		r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] \; ,
	\label{eq:reward2}\end{equation}
	where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is a proxy of the residual option price defined by
	\begin{equation}
	\begin{aligned}
		V_{\text{BS}}(T_k) &\coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}},K),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \; ,\\
		V_{\text{BS}}(T_0)&=0 \; , 
	\end{aligned}
	\end{equation} 
	with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} evaluated at the state $s_k$. In this form the agent actions are hidden inside the term $I_t$ used to compute the TVS forward curve $F^{\text{TVS}}$ defined by Equation~\eqref{eq:tvs_forward}. In this case, the reward function does not suffer of nil values for $0<T_k<T$: the RL agent always gets a feedback from the environment for its choices.  Notice that the hyper-parameter $\gamma$ plays the role of a discount factor in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future.
	If we take the cumulative reward per episode and set the PPO parameters\footnote{We refer to \cite{Schulman2016} for a more detailed description for the generalized advantage estimation parameter $\lambda$.} $\gamma =\lambda =1$ we obtain
	\begin{equation}
		R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma=\lambda =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+ \; ,
	\end{equation} 
	which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we conjecture that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

	In the \textsc{OpenAI Baselines} implementation of the PPO, the agent policy is parameterized again by a neural network; as for the previous method we have chosen an FFNN. More precisely, training is based on a stochastic policy, i.e. the action is drawn from a probability law, and the neural network is used to determine its parameters. In settings like ours where the action has a continuous domain, such law is a multivariate diagonal Gaussian distribution where the mean $\mu^\theta(s_k)$ is the output vector of the FFNN and the log-standard deviation $\log\sigma$ is an additional trainable parameter:
	\begin{equation}
		\pi^\theta(s_k) \sim \mathcal{N}(\mu^\theta(s_k),e^{\log\sigma}) \; .
	\end{equation}
	As one can observe, $\log\sigma$ is state-independent, but it is reduced as the number of the PPO update iterations increases. The idea is that the log-standard deviation will be higher at the beginning of the training phase in order to guarantee a good exploration of the action space while it will be lower at the end to avoid too much noise in the proximity of the optimal policy. Eventually at the end of training, the standard deviation is put to zero, so that the action becomes deterministically equal to the output of the FFNN.

@@

	\section{Numerical investigations}\label{sec:Numerical_investigations}
	Here we present the numerical results obtained with our proposed methods. We focus our analysis on a European call option on a TVS with the following product details
	\begin{equation}\nonumber
		I_0 = K = 1 \, \text{EUR} \; , \quad T=2 \, \text{yr} \; , \quad \bar{\sigma} = 5\% \; ,
	\end{equation}
	where $I_0$ is the starting value of the TVS-based portfolio.
	Moreover, without loss of generality, we consider the case of an unconstrained allocation strategy $\alpha$. The extension to the constrained case is easy.

@@

	Although in Section~\ref{subsec:BS_model} we have proved that under the Black and Scholes model the Equation~\eqref{eq:BS_optimal_strategy} solves the control problem, we want to take advantage of this \textit{a priori} solution as a benchmark to gather evidence on the robustness of our RL approach and to check if our analytical result is correct. Moreover, we use the BS model as numerical laboratory to perform fine-tuning tests for the RL algorithms hyper-parameters and analyze how they impact the final results and performances.

	We recall that in both the algorithms we parameterize the agent policy with an FFNN; thus this is completely characterized by the following hyper-parameters: number of hidden layers, number of neurons per hidden layer, and activation function per hidden layer. This is due to the fact that in this RL problem the number of the input neurons is equal to the state space dimension \eqref{eq:state}, while the number of the output ones coincides with the action space dimension \eqref{eq:action}. It is well known in the literature that neural networks give better performances in the training phase if the input data are well normalized \cite{Puheim2014}, \cite{Sola1997}. Thus we choose as state $s_k$ the following normalized block  
	\begin{equation}
		[\log(S_{T_k}/F(0,T_k)), I_{T_k}/I_0, T_k ] \quad \forall T_k\in\mathcal{T} \; ,
	\label{eq:input_state}\end{equation}
	where $F(t,T)$ is the forward curve vector of the risky assets from $t$ to $T$. In Equation~\eqref{eq:input_state} we have chosen as Markovian state $X_{T_k}$ the martingale term of the securities dynamics. In this way, we have that in the input block the variables have the same order of magnitude.

	\subsection{Black and Scholes: hyper-parameters fine tuning}
	We use the BS environment as a toy model to understand which parameters of the RL algorithm play key roles in the training and testing phase. 
	Our first approach has been the direct policy one since it represents a natural way to tackle the problem: since our goal is to find the allocation 
	strategy that maximizes the option price, we update the FFNN weights following the gradient direction of the loss function defined in Equation~\eqref{eq:direct_policy_loss}.

	We try to investigate the following hyper-parameters: the FFNN architecture, in particular which feature between the depth and the width of the network is more important, the activation function, and the learning rate of the optimizer. Moreover, we compare the performances of two well-known optimizers in Machine Learning literature: Nadam \cite{Dozat2016} and RMSprop \cite{Hiton2012}. Since we deal with a free allocation strategy $\alpha$ that can assume negative values, we have chosen among the wide variety of activation functions\cite{Goodfellow2016} the tanh and the elu. In this way, we can analyze the performance of a saturating activation function and a non-saturating one. Firstly we have performed a grid search on the learning rate starting value and we have found $10^{-3}$ a good choice in terms of speed of learning and avoiding over-fitting. Note that to detect over-fitting one typically compares the performance on one or more training sets with that on corresponding validation sets. Since our environment is simulated, there is no need to explicitly set aside training data for such task, nor to perform $K$-fold validation: indeed we can always generate for the purpose a set of new scenarios which were not seen during optimization.
	\begin{figure}[h!]
		\centering
		\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_tanh_grey.eps}
		\includegraphics[width=3.4in]{BS_fine_tuning_Nadam_elu_grey.eps}
		\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes scenario. The solid lines are the loss functions defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the Nadam with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
		\label{fig:result_bs_nadam}
	\end{figure}  
	\begin{figure}[h!]
		\centering
		\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_tanh_grey.eps}
		\includegraphics[width=3.4in]{BS_fine_tuning_RMSprop_elu_grey.eps}
		\caption{Learning curves of the direct policy algorithm applied to the Black and Scholes scenario. The solid lines are the loss values defined in Equation~\eqref{eq:direct_policy_loss} in function of the number of training epochs. Each continuous line represents a different neural network architecture with tanh activation function (top) and elu activation function (bottom). The optimizer adopted is the RMSprop with a learning rate of $10^{-3}$. The horizontal grey-dashed line is the conservative option price obtained by maximizing the TVS drift through Equation~\eqref{eq:analytical_solution_free_strategy}}
		\label{fig:result_bs_rmsprop}
	\end{figure}

	In Figures~\ref{fig:result_bs_nadam} and \ref{fig:result_bs_rmsprop} we present the learning curves of our fine-tuning tests for RMSProp and Nadam respectively. The three lines of each plot display the learning curves for different FFNN architectures: a one hidden layer network with 50 neurons, a 2 hidden layers with 30 and 20 neurons respectively, and a deeper one with three layers with 20, 15, and 5 neurons. Each learning curve is compared with the optimal option price (grey-dashed line) that we have computed with the closed form solution we have derived for the BS model in Section~\ref{subsec:BS_model}. We can see that all the learning curves converge to our expected price, providing a numerical finding of our theoretical result. More deeply, from the plots we observe that the RMSprop optimizer outperforms the Nadam. Moreover, the tanh activation function seems to be more preferable than the non-vanishing elu. However, more importantly, we have evidence that a deeper architecture of the neural network outperforms the shallow one. All the learning curves we have presented are the best-in-sample results, in terms of performance, of four runs with different random starting guesses for the hidden weights $\theta$. This procedure is necessary since the objective function is not convex.

	We take the $20\times 15 \times 5$ network with tanh from RMSprop as the best optimized network, and we run a Monte Carlo simulation with $10^6$ never-seen scenarios to check if the agent overfits the new data. 
	\begin{table}
	\begin{center}
	\caption{Comparison of TVO prices under Black and Scholes scenario: analytical solution price and direct policy reinforcement learning. The option parameters are: $I_0=K=1\,[\text{EUR}]$, $T=2\,[\text{yr}]$ and $\bar{\sigma}=5\%$}
	\label{tab:mc_results_bs}
	\begin{tabular}{ll}
	\hline\noalign{\smallskip}
	Pricing method & TVO price [EUR] \\
	\noalign{\smallskip}
	\hline
	\noalign{\smallskip}
	Analytical Solution & $4.2634\times 10^{-2}$ \\
	Direct policy RL & $(4.2624 \pm 0.005) \times 10^{-2}$ \\ 
	\hline
	\end{tabular}
	\end{center}
	\end{table}
	\setlength{\tabcolsep}{1.4pt}
	We report the results in Table~\ref{tab:mc_results_bs}: the RL price is compatible with the closed formula price. Thus the RL agent did not overfit the data during the training phase.
